---
title: "Clean Table S1 content and save it in RData object"
author: "Bea Szeitz"
output: html_notebook:
  toc: yes
---


```{r include=FALSE}
source("Load_packages.R")
source("Utility_functions.R")
```

What to run?

```{r }
get_broad_genesets <- F # takes ca. 1 min
run_singscore <- F # takes ca. 2 min

```


# Save genesets in a format that can be used for overrepresentation analysis (ORA) or pre-ranked GSEA


File requirement: 
- "Input_files/msigdb_v7.5.1_files_to_download_locally/msigdb_v7.5_GMTs/msigdb.v7.5.symbols.gmt"
- "Input_files/msigdb_v7.5_files_to_download_locally/msigdb_v7.5_GMTs/msigdb.v7.5.symbols.gmt"

Donwloadable from https://data.broadinstitute.org/gsea-msigdb/msigdb/release/7.5.1/ (last accessed on 02 June 2023)


```{r }
genesets_for_ora <- qusage::read.gmt("Input_files/msigdb_v7.5.1_files_to_download_locally/msigdb_v7.5.1_GMTs/msigdb.v7.5.1.symbols.gmt")
genesets_for_ora <- genesets_for_ora[grep("HALLMARK_|REACTOME_|KEGG_", names(genesets_for_ora))]

genesets_for_gsea<- clusterProfiler::read.gmt("Input_files/msigdb_v7.5.1_files_to_download_locally/msigdb_v7.5.1_GMTs/msigdb.v7.5.1.symbols.gmt")
genesets_for_gsea <- genesets_for_gsea[grepl("HALLMARK_|KEGG_|REACTOME_",genesets_for_gsea$term),]

save(genesets_for_ora, genesets_for_gsea, file="RData/HM_KEGG_Reactome_genesets.RData")

```




# Load the content of Table S1


```{r }

pROIs_lfq <- read_gct_13("Input_files/TableS1_ROI_annotation_expression_table.xlsx", "pROI_LFQ_table")
pROIs_ibaq <- read_gct_13("Input_files/TableS1_ROI_annotation_expression_table.xlsx", "pROI_iBAQ_table")
tROIs_beforelog2 <- read_gct_13("Input_files/TableS1_ROI_annotation_expression_table.xlsx", "tROI_table")
tROIs <- read_gct_13("Input_files/TableS1_ROI_annotation_expression_table.xlsx", "tROI_table_log2")

pROIs_lfq$Sample_annot$TIL <- as.numeric(pROIs_lfq$Sample_annot$TIL)
pROIs_lfq$Sample_annot$Mucin <- as.numeric(pROIs_lfq$Sample_annot$Mucin)
pROIs_lfq$Sample_annot$Stroma <- as.numeric(pROIs_lfq$Sample_annot$Stroma)
pROIs_lfq$Sample_annot$Immune_score_avr <- as.numeric(pROIs_lfq$Sample_annot$Immune_score_avr)
pROIs_lfq$Sample_annot$OS_years <- as.numeric(pROIs_lfq$Sample_annot$OS_years)

tROIs$Sample_annot$Immune_score <- as.numeric(tROIs$Sample_annot$Immune_score)
tROIs$Sample_annot$OS_years <- as.numeric(tROIs$Sample_annot$OS_years)

```


# Perform filtering and imputation in pROI tables

```{r }
pROIs_lfq$Expr_filt <- filter_missingvalues(pROIs_lfq$Expr, 80)

pROIs_ibaq$Expr_genenames <- pROIs_ibaq$Expr[pROIs_ibaq$Row_annot[!is.na(pROIs_ibaq$Row_annot$Gene),"Row_name"],]
row.names(pROIs_ibaq$Expr_genenames) <- pROIs_ibaq$Row_annot[!is.na(pROIs_ibaq$Row_annot$Gene),"Gene"]
pROIs_ibaq$Expr_genenames_filt <- filter_missingvalues(pROIs_ibaq$Expr_genenames, 80)

set.seed(12345)
pROIs_lfq$Expr_filt_imp <- impute.MinProb(pROIs_lfq$Expr_filt)
set.seed(12345)
pROIs_ibaq$Expr_genenames_filt_imp <- impute.MinProb(pROIs_ibaq$Expr_genenames_filt)

```


# SingScore calculation with iBAQ values and gene counts


File requirement: "Input_files/msigdb_v7.5.1_files_to_download_locally/msigdb_v7.5.1.xml"

Donwloadable from https://www.gsea-msigdb.org/gsea/downloads_archive.jsp (last accessed on 02 June 2023)


```{r }

if (get_broad_genesets){
  gsc <- getBroadSets("Input_files/msigdb_v7.5.1_files_to_download_locally/msigdb_v7.5.1.xml")
  save(gsc, file="RData/msigdb_v7.5.1.RData")
}

load("RData/msigdb_v7.5.1.RData") # saved as RData object for quicker import
genesets_to_use <- gsc[grepl("HALLMARK_|REACTOME_|KEGG_",names(gsc)) ]

```


```{r }
dgelists <- list(tROIs = DGEList(counts = tROIs$Expr, 
                               genes = row.names(tROIs$Expr)),
                 pROIs = DGEList(counts = pROIs_ibaq$Expr_genenames_filt_imp, 
                                genes = row.names(pROIs_ibaq$Expr_genenames_filt_imp)))

dgelists_ranked <- lapply(dgelists, rankGenes)
```


```{r  }
if (run_singscore){
  starttime <- Sys.time()
  SingScore_res <- lapply(dgelists_ranked, function(df){
    calculate_singScores(ranked_df = df, 
                         geneset_list = genesets_to_use, 
                         min_overlap = 10)
  })
  # 2 min runtime for the HM, KEGG, Reactome genesets for the 2 datasets
  endtime <- Sys.time()
  save(SingScore_res, file="RData/SingScores_KEGG_HM_Reac.RData")
}
load("RData/SingScores_KEGG_HM_Reac.RData")

```



```{r }
SingScore_res <- lapply(SingScore_res, function(x){
  lapply(x, function(y){as.data.frame(t(y))})
})

```


Create tROI singScore/ gene count table where ROIs from the same pROIs are summed up


```{r }

#pROI_samples_with_tROI <- na.omit(unique(tROIs$Sample_annot$Proteomic_region))
#matrix_to_use <- SingScore_res$tROIs$Scores
#tROI_annot <- tROIs$Sample_annot
#annot_col <- "Proteomic_region"

sum_up_tROI_to_pROI <- function(pROI_samples_with_tROI, matrix_to_use, tROI_annot, annot_col){
  
  summed_tab <- as.data.frame(matrix(nrow=nrow(matrix_to_use),
                                             ncol=length(pROI_samples_with_tROI)))
  row.names(summed_tab) <- row.names(matrix_to_use)
  colnames(summed_tab) <- pROI_samples_with_tROI
  
  
  for (i in 1:nrow(summed_tab)){
    for (j in 1:ncol(summed_tab)){
      original_ROIs <- tROI_annot[tROI_annot[,annot_col] == colnames(summed_tab)[j] &
                                            !is.na(tROI_annot[,annot_col]),"Sample"]
      original_values <- as.numeric(matrix_to_use[row.names(summed_tab)[i],
                                                               original_ROIs])
      summed_tab[i,j] <- mean(original_values, na.rm=T)
    }
  }
  
  colnames(summed_tab) <- pROI_samples_with_tROI
  return(summed_tab)
}

tROIs_sum <- tROIs
tROIs_sum$Expr <- sum_up_tROI_to_pROI(pROI_samples_with_tROI = na.omit(unique(tROIs$Sample_annot$Proteomic_region)),
                                      matrix_to_use = tROIs_sum$Expr,
                                      tROI_annot = tROIs$Sample_annot,
                                      annot_col = "Proteomic_region"
                                      )
tROIs_sum$Sample_annot <- pROIs_lfq$Sample_annot[colnames(tROIs_sum$Expr),]

tROIs_sum$Singscores <- filter_missingvalues(sum_up_tROI_to_pROI(pROI_samples_with_tROI = na.omit(unique(tROIs$Sample_annot$Proteomic_region)),
                                      matrix_to_use = SingScore_res$tROIs$Scores,
                                      tROI_annot = tROIs$Sample_annot,
                                      annot_col = "Proteomic_region"
                                      ),1)


tROIs$Singscores <- filter_missingvalues(SingScore_res$tROIs$Scores,1)
pROIs_lfq$Singscores <- filter_missingvalues(SingScore_res$pROIs$Scores,1)

```


# SafeTME estimations


```{r }
per_observation_mean_neg <- as.matrix(tROIs_beforelog2$Expr)["Negative Probe", ] # estimate per-observation background with negative probe
bg <- sweep(as.matrix(tROIs_beforelog2$Expr) * 0, 2, per_observation_mean_neg, "+") # define a background matrix

#nuclei_count <- as.numeric(tROIs_beforelog2$Sample_annot$AOI_nuclei_count)
#names(nuclei_count) <- tROIs_beforelog2$Sample_annot$Sample

```


Load the cell profile matrix (SafeTME)

```{r }
data("safeTME") # reached through the spatialdecon R package

```

Performing deconvolution with the spatialdecon function

```{r }

spatialDecon_res <- spatialdecon(norm = as.matrix(tROIs_beforelog2$Expr),
                                 bg = bg,
                                 X = safeTME,
                                 #cell_counts = nuclei_count[colnames(tROIs_beforelog2$Expr)], 
                                 ## including nuclei counts did not change betas
                                 align_genes = TRUE)


```


```{r }

tROIs$TME_estimation <- as.data.frame(spatialDecon_res$beta)
tROIs_sum$TME_estimation <- sum_up_tROI_to_pROI(pROI_samples_with_tROI = na.omit(unique(tROIs$Sample_annot$Proteomic_region)),
                                      matrix_to_use = tROIs$TME_estimation,
                                      tROI_annot = tROIs$Sample_annot,
                                      annot_col = "Proteomic_region"
                                      )

#save(TME_estimation, file="RData/TME_estimation.RData")

```



# Scaled and long tables


```{r }
tROIs$Expr_scaled <- as.data.frame(t(scale(t(tROIs$Expr))))

tROIs_sum$Expr_scaled <- sum_up_tROI_to_pROI(pROI_samples_with_tROI = na.omit(unique(tROIs$Sample_annot$Proteomic_region)),
                                      matrix_to_use = tROIs$Expr_scaled,
                                      tROI_annot = tROIs$Sample_annot,
                                      annot_col = "Proteomic_region"
                                      )

#pROIs_lfq$Expr_scaled <- as.data.frame(t(scale(t(pROIs_lfq$Expr))))
#pROIs_lfq$Expr_filt_scaled <- as.data.frame(t(scale(t(pROIs_lfq$Expr_filt))))
pROIs_lfq$Expr_filt_imp_scaled <- as.data.frame(t(scale(t(pROIs_lfq$Expr_filt_imp))))

```



```{r }

pROIs_lfq$Long_expr_filt_imp <- create_long_table(table_to_melt = pROIs_lfq$Expr_filt_imp,
                                             #genes = FALSE, 
                                             row_annot <- pROIs_lfq$Row_annot,
                                             col_annot <- pROIs_lfq$Sample_annot)

pROIs_lfq$Long_expr_filt_imp_scaled <- create_long_table(table_to_melt = pROIs_lfq$Expr_filt_imp_scaled,
                                                    #genes = FALSE, 
                                                    row_annot <- pROIs_lfq$Row_annot,
                                                    col_annot <- pROIs_lfq$Sample_annot)


tROIs$Long_expr <- create_long_table(table_to_melt = tROIs$Expr,
                                             #genes = TRUE, 
                                             row_annot <- tROIs$Row_annot,
                                             col_annot <- tROIs$Sample_annot)
tROIs$Long_expr_scaled <- create_long_table(table_to_melt = tROIs$Expr_scaled,
                                             #genes = TRUE, 
                                             row_annot <- tROIs$Row_annot,
                                             col_annot <- tROIs$Sample_annot)


tROIs_sum$Long_expr <- create_long_table(table_to_melt = tROIs_sum$Expr,
                                             #genes = TRUE, 
                                             row_annot <- tROIs_sum$Row_annot,
                                             col_annot <- tROIs_sum$Sample_annot)
tROIs_sum$Long_expr_scaled <- create_long_table(table_to_melt = tROIs_sum$Expr_scaled,
                                             #genes = TRUE, 
                                             row_annot <- tROIs_sum$Row_annot,
                                             col_annot <- tROIs_sum$Sample_annot)


```


```{r }
save(pROIs_lfq, tROIs, tROIs_sum, file="RData/Cleaned_input_files.RData")
```



```{r }

```




